---
title: "Assignment 8: Time Series Analysis"
author: "Connie Xiong"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics (ENV872L) on time series analysis.

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Use the lesson as a guide. It contains code that can be modified to complete the assignment.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
Space for your answers is provided in this document and is indicated by the ">" character.
If you need a second paragraph be sure to start the first line with ">".
You should notice that the answer is highlighted in green by RStudio. 
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
You will need to have the correct software installed to do this (see Software Installation Guide)
Press the `Knit` button in the RStudio scripting panel.
This will save the PDF output in your Assignments folder.
8. After Knitting, please submit the completed exercise (PDF file) to the dropbox in Sakai. Please add your last name into the file name (e.g., "Salk_A08_TimeSeries.pdf") prior to submission.

The completed exercise is due on Tuesday, 19 March, 2019 before class begins.

## Brainstorm a project topic
1. Spend 15 minutes brainstorming ideas for a project topic, and look for a dataset if you are choosing your own rather than using a class dataset. Remember your topic choices are due by the end of March, and you should post your choice ASAP to the forum on Sakai.

Question: Did you do this?

> ANSWER:  Yes, I have posted my topic choice to the forum.

## Set up your session 
2. Set up your session. Upload the EPA air quality raw dataset for PM2.5 in 2018, and the processed NTL-LTER dataset for nutrients in Peter and Paul lakes. Build a ggplot theme and set it as your default theme. Make sure date variables are set to a date format.

```{r}

getwd()
library(tidyverse)
library(dplyr)
library(trend)
library(lubridate)
library(nlme)
library(lsmeans)
library(multcompView)

#Import PM 2.5 Data for 2018
pm25_2018 <- read.csv("./Data/Raw/EPAair_PM25_NC2018_raw.csv")
#Set Date as date format
pm25_2018$Date <- as.Date(pm25_2018$Date,format = "%m/%d/%y")

#Import Peter Lake and Paul Lake for the nutrient data
PeterPaul.chem.nutrients <- read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv")
#Set sample date as date format
PeterPaul.chem.nutrients$sampledate <- as.Date(PeterPaul.chem.nutrients$sampledate, format = "%Y-%m-%d")

#Build a ggplot theme and set it to default
mytheme <- theme_classic() +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "right")
theme_set(mytheme)

```


## Run a hierarchical (mixed-effects) model

Research question: Do PM2.5 concentrations have a significant trend in 2018?

3. Run a repeated measures ANOVA, with PM2.5 concentrations as the response, Date as a fixed effect, and Site.Name as a random effect. This will allow us to extrapolate PM2.5 concentrations across North Carolina.

3a. Illustrate PM2.5 concentrations by date. Do not split aesthetics by site.

```{r}
pm25repeatedANOVA <- lme(data = pm25_2018,
                 Daily.Mean.PM2.5.Concentration ~ Date,
                 random = ~1|Site.Name)

summary(pm25repeatedANOVA)

#A rouph plot of the pm2.5 by date
plotpm25 <- ggplot(pm25_2018, aes(x= Date, y=Daily.Mean.PM2.5.Concentration))+
  geom_point()
print(plotpm25)

```

3b. Insert the following line of code into your R chunk. This will eliminate duplicate measurements on single dates for each site.
PM2.5 = PM2.5[order(PM2.5[,'Date'],-PM2.5[,'Site.ID']),]
PM2.5 = PM2.5[!duplicated(PM2.5$Date),]

3c. Determine the temporal autocorrelation in your model. 

3d. Run a mixed effects model. 

```{r}

# Eliminate duplicate measurements on single dates for each site
pm25_2018 = pm25_2018[order(pm25_2018[,'Date'],-pm25_2018[,'Site.ID']),]
pm25_2018 = pm25_2018[!duplicated(pm25_2018$Date),]

pm25mixed <- lme(data = pm25_2018,
                 Daily.Mean.PM2.5.Concentration ~ Date,
                 random = ~1|Site.Name,
                  correlation = corAR1(form = ~ Date|Site.Name),
                  method = "REML")

summary(pm25mixed)

```

Is there a significant increasing or decreasing trend in PM2.5 concentrations in 2018? 

> ANSWER: 

3e. Run a fixed effects model with Date as the only explanatory variable. Then test whether the mixed effects model is a better fit than the fixed effect model. 

```{r}
#Run a fixed effects model with Date as the only explanatory variable

pm25fixed <- gls(data = pm25_2018,
                 Daily.Mean.PM2.5.Concentration ~ Date,
                 method = "REML")
   
summary(pm25fixed)  

anova(pm25mixed, pm25fixed)

```


Which model is better?

> ANSWER: The mixed model is better because it has a lower AIC and did a better job in capturing the variability in the data.


## Run a Mann-Kendall test

Research question: Is there a trend in total N surface concentrations in Peter and Paul lakes? 

4. Duplicate the Mann-Kendall test we ran for total P in class, this time with total N for both lakes. Make sure to run a test for changepoints in the datasets (and run a second one if a second change point is likely). 

```{r}
#Wrangle the data and eliminate NAs in total P
PeterPaul.nutrients <- 
  PeterPaul.chem.nutrients %>%
  filter(depth == 0) %>%
  filter(!is.na(tp_ug))

#Run a Mann-Kendall test to see if there is a trend
mk.test(PeterPaul.nutrients$tp_ug)

#Run a test to determine change point
pettitt.test(PeterPaul.nutrients$tp_ug)

#As the test result shows that at position 50, there's likely a change; so we run a Mann-Kendall test for the points leading up to 50, which shows that there is a negative trend, but it is not significant; then run a test from 50 to the end, the result shows that there is a negative trend and it is statistically significant.
mk.test(PeterPaul.nutrients$tp_ug[1:49])
mk.test(PeterPaul.nutrients$tp_ug[50:263])

# Run a test to see if there is a second changing point
pettitt.test(PeterPaul.nutrients$tp_ug[50:263])
#The second changing point seems to occur at 50+161 = 211

mk.test(PeterPaul.nutrients$tp_ug[50:210])
mk.test(PeterPaul.nutrients$tp_ug[211:263])
#The result shows that both sections don't have a signficant trends

```


What are the results of this test?

> ANSWER: There isn't an overall trend in the total nitrogen in both lakes across the whole sample date. The result of the test shows that at position 50, which is Februry 2nd, 1992, there is a changing point in the trend of PM2.5 concentration(p<0.001). Before this date, there is no signidicant trend; after this date, there is a significant negative trend(z=-2.5, p=0.01). Then there is a second changing point at position 211, which is May 25th, 1998(p<0.005). Before and after this date, the trends are positive, but not significant.

5. Generate a graph that illustrates the TN concentrations over time, coloring by lake and adding vertical line(s) representing changepoint(s).

```{r}

plotTN <- ggplot(data = PeterPaul.nutrients, aes(x = sampledate, y = tp_ug, color = lakename)) +
  geom_point() +
  xlab(expression("Sample Date")) +
  ylab(expression(paste("Total Nitrogen Concentration (",mu,"g/L)"))) +
  labs(color="Lake Name") +
  scale_color_manual(values = c("blue", "purple")) +
  geom_vline(xintercept=as.Date("1992-02-02"), lty=5) +
  geom_vline(xintercept=as.Date("1998-05-25"), lty=5)
print(plotTN)

 ```